Cognitive processes are robust to early environmental conditions in two lizard species

### 2025-03-30 ms.qmd
Updates
Updates EcoEvoRxiv
updates

---
title: "Cognitive processes are robust to early environmental conditions in two lizard species"
format:
  docx: 
    number_sections: true
    linestretch: 2
    fontsize: 12pt
    mainfont: Times New Roman
    bibliography: "./bib/refs.bib"
    csl: "./bib/behavioral-ecology.csl"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

Pablo Recio ORCID: 0000-0002-5890-0218  
Dalton C. Leibold ORCID: 0000-0001-9645-2033  
Ondi L. Crino ORCID: 0000-0001-5700-1387  
Kristoffer H. Wild ORCID: 0000-0001-6714-3311  
Christopher R. Friesen ORCID: 0000-0001-5338-7454  
Basile Mauclaire ORCID:  
Amelia Y. Peardon ORCID: 0009-0005-6227-8379  
Daniel W.A. Noble ORCID: 0000-0001-9460-8743

@@

Here, we investigated the combined effects of prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature on cognition in two species of skinks, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). We manipulated CORT levels in the eggs and then incubated them at two different temperatures in a 2X2 factorial design. Post-incubation, juveniles were subjected to a color-associative task to assess their learning abilities. We hypothesized that prenatal CORT levels and thermal environment would impact the learning of an association task. We predicted that individuals exposed to high levels of CORT or low temperatures would learn slower than control individuals or those incubated at higher temperatures [@zhu_prenatal_2004; @amiel_hotter_2012; @eaton2015mild; @amiel_effects_2017]. Additionally, we predicted that incubation at high temperatures would mitigate the impact of CORT on learning performance, while cold incubation temperatures were expected to enhance the effects of CORT. The interactive effects of CORT and temperature may occur for two reasons that are not mutually exclusive: first, an increase in temperature leads to an overall rise in neural density [@amiel_effects_2017], thereby counteracting the impact of CORT [see @lemaire_prenatal_2000; @zhu_prenatal_2004; @eaton2015mild]; and second, the elevated metabolic rate associated with higher temperatures could accelerate CORT metabolism, resulting in embryos being exposed to CORT for a shorter time.

@@

*Eggs collection and incubation* -- Between mid-October 2022 to the end of February 2023, we provided females with a place to lay eggs by means of small boxes (12.5 L x 8.3 W x 5 H cm) containing moist vermiculite. These boxes were placed in one side of the communal enclosures (see above). We checked for eggs in the boxes three days a week (Monday, Wednesday, and Friday). After collection, we measured the length and width of eggs with a digital caliper to the nearest 0.1 mm, and mass using a digital scale (OHAUS, Model spx123; ± 0.001 g error). Then, we assigned individual IDs to the clutch and each egg. Eggs were treated with a CORT or control treatment (see CORT and Temperature Manipulation below) and were placed in individual cups (80 mL) with moist vermiculite (12 parts water to 4 parts vermiculite). The cups were covered with cling wrap to retain moisture and left in LATWIT 2X5D-R1160 incubators at two different temperatures (see CORT and Temperature manipulation below) until hatching. Eggs were then checked three times a week for hatchlings.

@@

#### CORT and Temperature Manipulation  
To empirically test the effect of early environment, we manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (‘Cold’ - 23 ± 3 ºC or ‘Hot’ - 28 ± 3 ºC) in a 2x2 factorial design (@fig-Methods A). We topically dosed eggs with a CORT solution (10 pg/mL) or a control treatment (100% ethanol). Corticosterone treatments were made by dissolving crystalline corticosterone (Sigma, Cat. No. C2505) in 100% ethanol. To dose eggs, we applied 5µl of solutions to eggshells using a micropipette. We selected these doses based on previous studies publishing yolk CORT concentrations in other oviparous reptiles [@lovern2008effects; @hanover2019corticosterone], while also validating that it fell within the range of CORT concentrations in eggs within our population [@crino2024eggs]. CORT treatment increased mean yolk CORT levels ~3.7x higher than control eggs [@crino2024eggs]. After the treatment, the eggs were incubated in one of the two previously mentioned temperature regimes (‘Cold’ or ‘Hot’) until hatching. The number of eggs per clutch assigned to each hormone and temperature treatment was counterbalanced in both species.

@@

The experiment occurred between the 6^th^ of March to the 17^th^ of May 2023, and tests were performed between 1100 to 1200 when the lizards were active. Trials in the learning phase were recorded with CCTV systems, always using the same camera per individual. We recorded whether the animal chose the correct ramp first or not. We considered that a choice was made if the head of the lizard was inside the well of one of the ramps. We considered a trial to have failed if there was no choice after one hour of recording. These trials were scored as NA. We excluded from our analyses those individuals that performed inconsistently, as defined as not choosing in less than 20 out of 35 trials (~57%). We considered each lizard's first trial to be the first one where a choice was made.

@@

Alternatively, our observations could be obscured by other variables that are known to influence the extent and direction of the effects of GCs [@szuran_water_1994; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @farrell_developmental_2016]. GCs are known to have hormetic effects [@du_dynamic_2009]. Generally, low concentrations or short exposure to GCs have been shown to improve the rate at which animals learn while high concentrations or exposure to GCs during long periods have the opposite effects [@du_dynamic_2009; @mcewen_brain_2012]. Thus, high elevations in CORT could affect learning through permanent changes in brain function (programmatic effects), while lower doses can result in lifelong elevation of baseline CORT, affecting learning through activational effects. Our findings indicate that associative learning in _L. delicata_ and _L. guichenoti_ is robust to the programmatic effects of elevated CORT exposure. However, we cannot exclude the possibility that concentrations different from the one used here may affect learning abilities.

@@

#### Incubation temperature does not affect learning
We predicted hot-incubated lizards would perform better in the associative learning task, since most studies demonstrate enhanced learning abilities when eggs are incubated at higher temperatures [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014]. While the opposite pattern has been found in some studies, their incubation temperatures were far above the natural thermal range of the species in their hot treatments [@dayananda_incubation_2017; @abayarathna_effects_2020]. As such, it is unclear how such these extreme conditions relate to wild environments. In _L. delicata_ and _L. guichenoti_, associative learning appears robust to incubation temperature because of the lack of detectable effect temperature had on learning. Our results, therefore, do not align with previous studies. However, the effect of prenatal temperature on cognition, and brain physiology and structure has been investigated only in a small number of species [see @coomber_independent_1997; @sakata_neural_2000; @amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @dayananda_incubation_2017; @abayarathna_effects_2020], limiting our understanding on how early thermal environment can affect cognitive abilities. Our results suggest that the impact of incubation temperature on learning may not be as widespread across reptiles as we think.

@@

#### Learning rates between species did not vary  
We did not see any difference in learning rates when comparing both species. This result supports previous studies that found similar learning abilities in _L. delicata_ and _L. guichenoti_ [@bezzina2014does]. In the experiment conducted by @bezzina2014does, both species failed to complete the associative learning task under the authors' criterion, while in our experiment, both species completed the task, exhibiting similar learning rates. Complexity, experimental design, or the criterion employed to define learning could be the major cause of the discrepancies between ours and @bezzina2014does. Developing common strategies and approaches to assess animal learning could help understand how learning abilities are shaped in different taxa or environments.

>>>>

---
title: "Cognitive processes are robust to early environmental conditions in two lizard species"
docx: 
  number_sections: true
  linestretch: 2
  fontsize: 12pt
  mainfont: Times New Roman
bibliography: "./bib/refs.bib"
csl: "./bib/behavioral-ecology.csl"
execute:
  echo: false
  error: false
  cache: false
  warning: false
link-citations: true
crossref:  
  fig-title: Fig    # (default is "Figure")
  title-delim: —     # (default is ":")
  fig-prefix: Fig.   # (default is "Figure")
  tbl-prefix: Table  # (default is "Table")
editor_options: 
  chunk_output_type: console
---

@@

Pablo Recio ORCID: 0000-0002-5890-0218  
Dalton C. Leibold ORCID: 0000-0001-9645-2033  
Ondi L. Crino ORCID: 0000-0001-5700-1387  
Kristoffer H. Wild ORCID: 0000-0001-6714-3311  
Christopher R. Friesen ORCID: 0000-0001-5338-7454  
Basile Mauclaire ORCID: 0000-0003-3956-6055   
Amelia Y. Peardon ORCID: 0009-0005-6227-8379  
Daniel W.A. Noble ORCID: 0000-0001-9460-8743

@@

Here, we investigated the combined effects of prenatal corticosterone (CORT) - the main GC in reptiles - and incubation temperature on cognition in two species of skinks, the delicate skink (_Lampropholis delicata_) and the common garden skink (_L. guichenoti_). We manipulated CORT levels in the eggs and then incubated them at two different temperatures in a 2X2 factorial design. Post-incubation, juveniles were subjected to a color-associative task to assess their learning abilities. We hypothesized that prenatal CORT levels and thermal environment would impact the learning of an association task. We predicted that individuals exposed to high levels of CORT or low temperatures would learn slower than control individuals or those incubated at higher temperatures [@zhu_prenatal_2004; @amiel_hotter_2012; @eaton2015mild; @amiel_effects_2017]. Additionally, we predicted that incubation at high temperatures would mitigate the impact of CORT on learning performance, while cold incubation temperatures were expected to enhance the effects of CORT. The interactive effects of CORT and temperature may occur for two reasons that are not mutually exclusive: first, an increase in temperature leads to an overall rise in neural density [@amiel_effects_2017], thereby counteracting the impact of CORT [see @lemaire_prenatal_2000; @zhu_prenatal_2004; @eaton2015mild]; and second, the elevated metabolic rate associated with higher temperatures could accelerate CORT metabolism, resulting in embryos being exposed to CORT for a shorter time. In contrast, glucocorticoids in endotherms are associated with increased energy demands[e.g., @rubalcaba2022body], which could lead to higher CORT production in lizards incubated at warmer temperatures. However, previous research on _L. delicata_ [@crino2024eggs] has found that cold-incubated lizards had higher baseline CORT levels, suggesting that cooler incubation temperatures may increase the potential effects of CORT exposure.

@@

*Eggs collection and incubation* -- Between mid-October 2022 to the end of February 2023, we provided females with a place to lay eggs by means of small boxes (12.5 L x 8.3 W x 5 H cm) containing moist vermiculite. These boxes were placed in one side of the communal enclosures (see above). We checked for eggs in the boxes three days a week (Monday, Wednesday, and Friday). After collection, we measured the length and width of eggs with a digital caliper to the nearest 0.1 mm, and mass using a digital scale (OHAUS, Model spx123; ± 0.001 g error). Then, we assigned individual IDs to the clutch and each egg. Eggs were treated with a CORT or control treatment (see CORT and Temperature Manipulation below) and were placed in individual cups (80 mL) with moist vermiculite (12 parts water to 4 parts vermiculite). The cups were covered with cling wrap to retain moisture and left in LABWIT 2X5D-R1160 incubators at two different temperatures (see CORT and Temperature manipulation below) until hatching. Eggs were then checked three times a week for hatchlings.

@@

#### CORT and Temperature Manipulation  
To empirically test the effect of early environment, we manipulated CORT concentration in eggs and incubated them under one of two temperature regimes (‘Cold’ - 23 ± 3 ºC or ‘Hot’ - 28 ± 3 ºC) in a 2x2 factorial design (@fig-Methods A). We topically dosed eggs with a CORT solution (10 pg/mL) or a control treatment (100% ethanol). Corticosterone treatments were made by dissolving crystalline corticosterone (Sigma, Cat. No. C2505) in 100% ethanol. To dose eggs, we applied 5µl of solutions to eggshells using a micropipette. We selected these doses based on previous studies publishing yolk CORT concentrations in other oviparous reptiles [@lovern2008effects; @hanover2019corticosterone], while also validating that it fell within the range of CORT concentrations in eggs within our population [@crino2024eggs]. CORT treatment increased mean yolk CORT levels ~3.7x higher than control eggs [@crino2024eggs]. After the treatment, the eggs were incubated in one of the two previously mentioned temperature regimes (‘Cold’ or ‘Hot’) until hatching. These temperatures represent the upper and lower limit of the natural incubation temperatures [@qualls2000post; @cheetham2011embryonic]. The number of eggs per clutch assigned to each hormone and temperature treatment was counterbalanced in both species.

@@

The experiment occurred between the 6^th^ of March to the 17^th^ of May 2023, and tests were performed between 1100 to 1200 when the lizards were active. Trials in the learning phase were recorded with CCTV systems, always using the same camera per individual. All the videos were analysed by the same observer (PR) who was blind to the treatment of the lizards. We recorded whether the animal chose the correct ramp first or not. We considered that a choice was made if the head of the lizard was inside the well of one of the ramps. We considered a trial to have failed if there was no choice after one hour of recording. These trials were scored as NA. We excluded from our analyses those individuals that performed inconsistently, as defined as not choosing in less than 20 out of 35 trials (~57%). We considered each lizard's first trial to be the first one where a choice was made.

@@

Alternatively, our observations could be obscured by other variables that are known to influence the extent and direction of the effects of GCs [@szuran_water_1994; @crino_corticosterone_2014-learn; @farrell_developmental_2015-learn; @bebus_associative_2016; @farrell_developmental_2016]. GCs are known to have hormetic effects [@du_dynamic_2009]. Generally, low concentrations or short exposure to GCs have been shown to improve the rate at which animals learn while high concentrations or exposure to GCs during long periods have the opposite effects [@du_dynamic_2009; @mcewen_brain_2012]. Thus, high elevations in CORT could affect learning through permanent changes in brain function (programmatic effects), while lower doses can result in lifelong elevation of baseline CORT, affecting learning through activational effects. Our findings indicate that associative learning in _L. delicata_ and _L. guichenoti_ is robust to the programmatic effects of elevated CORT exposure. Similarly, exposing embryos at different stages of development could impact the effects of CORT. However, we cannot exclude the possibility that concentrations different from the one used here or exposure at different developmental times does affect learning.

@@

#### Incubation temperature does not affect learning
We predicted hot-incubated lizards would perform better in the associative learning task, since most studies in other species demonstrate enhanced learning abilities when eggs are incubated at higher temperatures [@amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014]. These studies employed incubation temperatures within natural nesting thermal limits. In contrast, in those studies where cold-incubated lizards outperformed hot incubated ones, the incubation temperatures employed in the hot condition were far above the natural thermal range of the species in their hot treatments [@dayananda_incubation_2017; @abayarathna_effects_2020]. As such, it is unclear how such these extreme conditions relate to wild environments. In _L. delicata_ and _L. guichenoti_, associative learning appears robust to incubation temperature because of the lack of detectable effect temperature had on learning. Our results, therefore, do not align with previous studies. However, the effect of prenatal temperature on cognition, and brain physiology and structure has been investigated only in a small number of species [see @coomber_independent_1997; @sakata_neural_2000; @amiel_hotter_2012; @amiel_egg_2014; @clark_colour_2014; @amiel_effects_2017; @dayananda_incubation_2017; @abayarathna_effects_2020], limiting our understanding on how early thermal environment can affect cognitive abilities. Our results suggest that the impact of incubation temperature on learning may not be as widespread across reptiles as we think.

@@

#### Learning rates between species did not vary  
We did not see any difference in learning rates when comparing both species. Since both species occupy similar habitats and have similar ecology [@chapple_know_2011; @chapple2013divergent; @chapple_biology_2014], it may not be surprising that cognitive abilities are similar between both species. Nonetheless, previous studies have shown that _L. guichenoti_ is less prone to explore novel environments than _L. delicata_, which may be related to the success of _L. delicata_ as an invasive species compared to _L. guichenoti_ [@chapple_know_2011]. Our results show that the ability of L. delicata to colonize new areas seems not to be related to learning abilities. Our findings support previous studies that found similar learning abilities in _L. delicata_ and _L. guichenoti_ [@bezzina2014does]. In the experiment conducted by @bezzina2014does, both species failed to complete the associative learning task under the authors' criterion, while in our experiment, both species completed the task, exhibiting similar learning rates. Complexity, experimental design, or the criterion employed to define learning could be the major cause of the discrepancies between ours and @bezzina2014does. Developing common strategies and approaches to assess animal learning could help understand how learning abilities are shaped in different taxa or environments.
